---
title: "House Gerrymandering Visualization"
author: "Jordan Isaacs (jisaacs)"
date: "10/16/2020"
output:
  html_document:
      toc:  true
      toc_float:  true
      code_folding:  show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message = F, warning = F}
library(tidyverse)
library(tidycensus)
library(reshape2)
library(gridExtra)
census_api_key("c056ebd4b54cffef27c51415099c17a03436ec92")
```

## Loading Data

* House election data is from [MIT Election Data + Science lab](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/IG0UN2#)
```{r, message = F, warning = F}
current_legislators <- read_csv("https://theunitedstates.io/congress-legislators/legislators-current.csv")
election_data.raw <- read_csv("./Data/1976-2018-house2.csv")

get_acs.us_districts <- function(years) {
  acs.us_districts <- list()
  i <- 1
  variables = c(population = "B01003_001",
                white_pop = "B02008_001",
                black_pop = "B02009_001",
                hisp_pop = "B03001_003",
                asian_pop = "B02011_001",
                native_pop = "B02010_001",
                median_age = "B01002_001",
                median_income_12mo = "B06011_001")
  
  for (year in years) {
    data <- get_acs(geography = "congressional district",
                    year = year,
                    variables = variables,
                    survey = "acs1")
    data <- list(data)
    acs.us_districts <- c(acs.us_districts, list(data))
  }
  
  
  names(acs.us_districts) <- years
  return (acs.us_districts)
}

years <- seq(2012, 2018, by = 2)
acs.us_districts.raw <- get_acs.us_districts(years)

# Minnesota uses/used affiliated parties
fix_mn = function(party) {
  if (is.na(party)) {
    return (party)
  }
  if (party == "democratic-farmer-labor") {
    return ("democrat")
  } else if (party == "independent-republican") {
    return ("republican")
  } else {
    return (party)
  }
}

election_data.raw$party = election_data.raw$party %>% sapply(fix_mn)
# Fix 0 district
election_data.raw = election_data.raw %>% mutate(district = ifelse(district == 0, 1, district))
```

## Process Election Data

+ Process election data to only have the general election data for each two year period (ignore special elections)
+ Aggregate up all democrat, republican, and other votes. (Even if multiple from same party were running)
+ Votes for a party take precedence over candidate (if candidate is democrat and third party, votes go to the party the voter chose)
```{r, message = F}
get_demvotes = function(party, candidatevotes) {
  return (sum(candidatevotes[party == "democrat"], na.rm = TRUE))
}

get_repubvotes = function(party, candidatevotes) {
  return (sum(candidatevotes[party == "republican"], na.rm = TRUE))
}

get_othervotes = function(party, candidatevotes) {
  return (sum(candidatevotes[!(party %in% c("democrat", "republican"))]))
}

election_data.processed <- election_data.raw %>%
  filter(special == FALSE, stage == "gen") %>%
  group_by(year, state_po, district, runoff) %>%
  summarise(democrat = get_demvotes(party, candidatevotes),
            republican = get_repubvotes(party, candidatevotes),
            other = get_othervotes(party, candidatevotes),
            total = max(totalvotes)) %>%
  arrange(year, state_po, district, desc(runoff)) %>%
  distinct(year, state_po, district, .keep_all = TRUE) %>%
  rename(runoff_ranked = runoff)

election_is_valid = function() {
  for (i in 1:nrow(election_data.processed)) {
    row = election_data.processed[i,]
    demvotes = as.numeric(row[["democrat"]])
    repubvotes = as.numeric(row[["republican"]])
    othervotes = as.numeric(row[["other"]])
    totalvotes = as.numeric(row[["total"]])
    if (demvotes + repubvotes + othervotes != totalvotes ||
        (demvotes == 0 && repubvotes == 0 && othervotes == 0)) {
      cat("Invalid row (",  i, "):\n", sep = "")
      print(as.character(row))
    }
  }
}

election_is_valid()
```

## Fix bad election data

+ Using data from [ballotpedia](https://ballotpedia.org/Main_Page)
+ For 2012 CT using [state election results](https://portal.ct.gov/-/media/SOTS/ElectionServices/ElectionResults/2012/2012ElectionResultspdf.pdf)
+ Maine votes are from final ranked choice
```{r}
# 6 missing other votes in this race
election_data.processed[5799, 7] = election_data.processed[5799, 7] + 6

# 1 vote extra in other and wrong total
election_data.processed[7900, 7] = election_data.processed[7900, 7] - 1
election_data.processed[7900, 8] = 73726 + 197163 + 20410

# Missing 1 write in vote
election_data.processed[8768, 7] = election_data.processed[8768, 7] + 1

# Messed up from ranked choice voting, using final round
election_data.processed[9319, 5] = 142440
election_data.processed[9319, 6] = 138931

# Setting runoff_ranked to true
election_data.processed[9319, 4] = TRUE

# Uncontested election that democrats one
election_data.processed[8797, 5] = 1
election_data.processed[8797, 8] = 1

election_is_valid()
```

## 2020 house data
+ Using Jonathan's Cervas scraped NYT data
```{r, message = F}
house_2020 <- read_csv("https://raw.githubusercontent.com/jcervas/2020-Elections/main/house2020.csv")
state_to_abbv <- function(state) {
  return (state.abb[state.name == state])
}

house_2020$state = sapply(house_2020$state, state_to_abbv)
house_2020$district = house_2020$district
house_2020 = house_2020 %>% select(state, district, dem, rep, other, totalvotes) %>%
  rename("democrat" = dem, "republican" = rep, "total" = totalvotes, "state_po" = state)
house_2020$runoff_ranked = NA
house_2020$year = 2020
election_data.processed = rbind(election_data.processed, house_2020)
```

## Stylistically alter election data

+ Some elections are uncontested or extremely partisan (over 75% for a party). In this case give 75% of votes to the winning party and 25% to the missing/losing party (either democrat or republican).
+ Drop third party voting data unless they have majority of vote share (won the election)
+ Note: By dropping third party voting data we will lose the ability to analyze voter turnout 

```{r}
alter_row <- function(x) {
  other = as.numeric(x[["other"]])
  democrat = as.numeric(x[["democrat"]])
  republican = as.numeric(x[["republican"]])
  total = as.numeric(x[["total"]])
  if (total == 0) {
    return (c(democrat, republican, other, total));
  }
  if (other > democrat && other > republican) {
    return (c(democrat, republican, other, total));
  } else {
    other = 0;
    total = democrat + republican;
    if (democrat / total > 0.75) {
      return (c(democrat * 0.75, democrat * 0.25, 0, total));
    } else if (republican / total > 0.75) {
      return (c(republican * 0.25, republican * 0.75, 0, total));
    }
    return (c(democrat, republican, 0, total));
  }
}

process_data <- function(data) {
  processed = data %>% apply(., 1, alter_row) %>% t %>% as.data.frame
  data$democrat = processed$V1
  data$republican = processed$V2
  data$other = processed$V3
  data$total = processed$V4
  return (data)
}

election_data.processed = process_data(election_data.processed)
```


- No third party has won since 2010. Mostly looking at data since 2010 so not going to research whether they caucused with a specific party.
```{r}
filter(election_data.processed, other != 0)
filter(election_data.processed, runoff_ranked == TRUE)
```




## Reshape Election Data

+ Turn from wide into long format
```{r}
election_data <- melt(data = election_data.processed,
                                id.vars = c("year", "state_po", "district"),
                                measure.vars = c("republican", "democrat", 
                                                 "other")) %>%
  rename(party = variable, votes = value)
```

## Process Census data

+ Process the NAME column into State and District
+ Add a year column
```{r}
process_census_data <- function(population_data, year, type) {
  population_data = population_data[[1]]
  state_matches = regexpr("(?<=, ).+", population_data$NAME, perl = TRUE)
  population_data$state <- regmatches(population_data$NAME, state_matches)
  
  district_matches <- regexpr("(?<=District )[[:digit:]]+", population_data$NAME, perl = TRUE)
  districts <- rep(0, length(district_matches))
  districts[district_matches != -1] <- regmatches(population_data$NAME, district_matches)
  population_data$district <- as.numeric(districts)
  population_data = population_data %>% mutate(district = ifelse(district == 0, 1, district))
  
  state_indexes <- match(tolower(population_data$state), tolower(state.name))
  population_data$state <- state.abb[state_indexes]
  population_data <- population_data %>%
    filter(state != is.na(state))
  
  if (type == "acs") {
    population_data <- population_data %>% select(state, district, variable, estimate) %>%
      pivot_wider(., id_cols = c("state", "district"), values_from = "estimate", names_from = "variable")
  } else if (type == "dec") {
    population_data <- population_data %>% select(state, district, value) %>%
      rename(district = districts, population = value)
  } else {
    return (NA)
  }
  
  population_data$year = rep.int(year, dim(population_data)[1])
  return (population_data)
}

acs.us_districts.processed = acs.us_districts.raw
for (year in years) {
  acs.us_districts.processed[[as.character(year)]] <- process_census_data(acs.us_districts.raw[[as.character(year)]], year, "acs")
}

```


## Join the data together

+ Row bind the population data then left join election data
+ Essentially adding voting data to our district population data

```{r}
join_data <- function(population_data, election_data) {
  combined_pop = population_data[[as.character(years[1])]]
  for (year in years) {
    if (year != years[1]) {
      combined_pop = bind_rows(combined_pop, population_data[[as.character(year)]])
    }
  }
  return (left_join(election_data, combined_pop, by = c("year" = "year",
                                               "state_po" = "state",
                                               "district" = "district")))
}
data <- join_data(acs.us_districts.processed,
                  election_data)
data = rename(data, "state" = state_po)
```


## Save as CSV

```{r}
write_csv(data, file="house_census_data.csv")
```

## Save as R object
```{r}
save(data, file = "house_census_data.Rdata")
```



